---
title: "Data Preprocessing: Cleaning and Profilingg"
author: "Elie Diwambuena"
date: "2023-04-24"
output: html_document
---


In this workboook, I am going to clean a public dataset on credit risk of customers 
of a typical bank published by SRIHARI P on Kaagle.

### Setting up R environment

As always, I start by loading the library and the dataset.

```{r}
library(tidyverse)

# load dataset
file_path = "credit_customers.csv"
data <- read.csv(file_path, stringsAsFactors = F)

```

Now that I have loaded the dataset and the required library, I will check for the
structure of the dataset and look for some disparities or inconsistencies.


### Familiarizing with the dataset

**check data structure**
```{r}
str(data)
```

As shown in the output, the data has 1000 observations and 21 variables. What is
quite noticeable are the weird symbols inside some variables such as checking_status,
employment, personal_status etc. This leads me to look into those variables and
try to understand if those annotations are errors or actual inputs. For that, I will
look closely into each variables that might carry potential errors.


**checking_status**

Let start by grouping data into checking_status groups or categories.

```{r}
data %>% group_by(checking_status) %>% summarise(count = n()) 
```

The presumed weird symbols are actually inputs. The variable "checking_satus" 
records whether the applicant or customer has a checking account and if so, the 
amount held in the account. Thus, we have 4 categories. The entry 0<=X<200 means that the customer has a checking account with an amount of less than 200. However, this way of recording inputs is very unprofessional and it makes difficult to interprete later on. Hence, I will label them differently, change the data type and order them. It makes sense to order these categories because a person with a negative balance in the checking
account will likely have higher chance of defaulting on a loan.


```{r}
data$checking_status <- factor(data$checking_status, 
                               levels = c("<0", "0<=X<200",">=200", "no checking"),
                               labels = c("negative", "less than 200", 
                                          "more than or = 200", "no checking"))
data %>% group_by(checking_status) %>% summarise(count = n()) 
```


î have labelled and ordered them as negative (<0), less than 200, more than or = 200,
and no checking. Then, I will do the same thing for the remaining variables.


**credit history**
``` {r}
data %>% group_by(credit_history) %>% summarise(count = n()) 
````

It had 5 categories in credit history. Let change the label and provide an order.
```{r}
data$credit_history <- factor(data$credit_history,
                              levels = c("critical/other existing credit",
                                         "delayed previously", "existing paid",
                                         "no credits/all paid", "all paid"),
                              labels = c("dully paid", "critical", "delayed",
                                        "dully paying", "no credits"))

data %>% group_by(credit_history) %>% summarise(count = n()) 
```


**savings status**

```{r}
data %>% group_by(savings_status) %>% summarise(count = n()) 
````

It also has 5 categories in savings status. let change the label and provide an order.

```{r}
data$savings_status <- factor(data$savings_status,
                              levels = c("no known savings", "<100", "100<=X<500",
                                         "500<=X<1000", ">=1000"),
                              labels = c("between 100 and 499", 
                                         "between 500 and 999",
                                         "less than 100",
                                         "more than or = 1000",
                                         "no savingss"))

data %>% group_by(savings_status) %>% summarise(count = n()) 
```


**employment: in years**

```{r}
data %>% group_by(employment) %>% summarise(count = n()) 
````

It has 5 categories in employment. I will again change the label and provide an order.

```{r}
data$employment <- factor(data$employment,
                              levels = c("unemployed", "<1", "1<=X<4",
                                         "4<=X<7", ">=7"),
                              labels = c("unemployed", "less than 1 year",
                                         "less than 4 years", "less than 7 years",
                                         "more than 7 years"))

data %>% group_by(employment) %>% summarise(count = n()) 
````


**purpose**

```{r}
data %>% group_by(purpose) %>% summarise(count = n()) 
```

It 10 categories in purpose. I am not going to rename them but I will change the 
data type to factor.

```{r}
data$purpose <- factor(data$purpose)
```


***foreign worker ***

```{r}
data %>% group_by(foreign_worker) %>% summarise(count = n()) 
````

It has 2 categories. Let change the data type to factor.

```{r}
data$foreign_worker <- factor(data$foreign_worker)
```


**own_telephone **

```{r}
data %>% group_by(own_telephone) %>% summarise(count = n()) 
````

It also has 2 categories. I will change the data type to factor and rename the label.

```{r}
data$own_telephone <- factor(data$own_telephone,
                             labels= c("no", "yes"))
````


**job**

```{r}
data %>% group_by(job) %>% summarise(count = n()) 
````

It has 4 categories. I will let change the label and the order.

```{r}
data$job <- factor(data$job,
                             levels = c("unemp/unskilled non res",
                                        "unskilled resident", "skilled",
                                        "high qualif/self emp/mgmt"),
                             labels= c("unskilled non-res", 
                                       "unskilled res",
                                       "skilled", 
                                       "highly qualified"))

data %>% group_by(job) %>% summarise(count = n()) 
```


**housing **

```{r}
data %>% group_by(housing) %>% summarise(count = n()) 
````

It has 3 categories.Again, let change the label and the order.

```{r}
data$housing <- factor(data$housing,
                   levels = c("rent", "for free", "own"),
                   labels= c("rented", "free use", "owned"))

data %>% group_by(housing) %>% summarise(count = n())
````
                         

**other payment plans: other debts **

```{r}
data %>% group_by(other_payment_plans) %>% summarise(count = n()) 
````

It has 3 categories. Let change the label and the order.

```{r}
data$other_payment_plans <- factor(data$other_payment_plans,
                       levels = c("bank", "stores", "none"),
                       labels= c("bank debt", "purchase debt", "no debt"))

data %>% group_by(other_payment_plans) %>% summarise(count = n()) 
````


**property magnitude: collateral**

```{r}
data %>% group_by(property_magnitude) %>% summarise(count = n()) 
```

It has 3 categories. Once again, let change the label and the order.

```{r}
data$property_magnitude <- factor(data$property_magnitude,
                                   levels = c("no known property", 
                                              "life insurance", 
                                              "car",
                                              "real estate"),
                                   labels= c("no collateral", 
                                             "life insurance", 
                                             "vehicle", 
                                             "real estate"))

data %>% group_by(property_magnitude) %>% summarise(count = n()) 
```


**other parties to the loan: guarantee **

```{r}
data %>% group_by(other_parties) %>% summarise(count = n()) 
````

It has 3 categories. I will change the label and the order.

```{r}
data$other_parties <- factor(data$other_parties,
                                  levels = c("none", 
                                             "co applicant",
                                             "guarantor"),
                                  labels= c("individual", 
                                            "co-aplicant", 
                                            "guaranteed"))

data %>% group_by(other_parties) %>% summarise(count = n()) 
````


**personal_status**

```{r}
data %>% group_by(personal_status) %>% summarise(count = n())  
````

What I notice is that this varibale containes two other variables: gender and civil status. I will therefore split it.

```{r}
# one for gender and the other for civil status

data <- data %>% 
  separate(personal_status, into = c("gender", "civil_status"), sep = " ")

````

Once I have splitted it, I can now rename the lables for the civil status variables and 
change their data types to factor.

```{r}
# factor gender
data$gender <- factor(data$gender)

# label civil status and order
data %>% group_by(civil_status) %>% summarise(count = n())  

data$civil_status <- factor(data$civil_status, 
                            levels = c("div/dep/mar","single", 
                                       "div/sep", "mar/wid"),
                            labels = c("unspecified", "single",
                                       "divorced", "married/widowed"))

````


**class :  this could be the prediction class**

```{r}
data %>% group_by(class) %>% summarise(count = n()) 
````

It has 2 classes. Let change the data type to factor.
```{r}
data$class <- factor(data$class)

````

At this point, I have familiarized myself with the data set. I know almost everything
about the dataset such as what types of variables it contains and what inputs those 
variables carry, the data types and the size of the dataset. 

It is not necessary and can be teadious to rename labels for each variables. This
can be later on a needed basis. However, I choose to rename them a this point as part
of my familiarizing process and also to make them convey more meaning. If I happen to
open this dataset after a long period, I will have no difficulties understanding it.



### Checking for missing values 

Now that I have checked the structure of my dataset and correct for datatypes and
labels, I will check if my dataset contains any missing values.

```{r}
any(is.na(data))
````

Fortunately, the dataset has no missing values at all. Therefore, I can conclude 
thereby the data preprocessing step. In the next workbook, I will apply some
visualization technics to the dataset to understand the relation between variables.


# End
